Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-16 sections/background_related.tex
Updates from Overleaf

Masuda \textit{et al.} also applied supervised learning~\cite{masuda2021soundmatch}. Their work highlights the issue of non-linearity in parameter-to-synthesizer outputs and out-of-domain search. This work uses a differentiable subtractive synthesizer (two oscillators and an LP filter) alongside a NN model pre-trained with P-Loss on an in-domain dataset of randomly selected parameters. After training, the model was fine-tuned using 20,000 out-of-domain sounds from the NSynth dataset~\cite{engel2017neural} and multi-scale spectrogram loss~\cite{engel2020ddsp}. This approach proved more effective---i.e, achieved lower multi-level spectral difference in out-of-domain tests---than baseline models, which were either not exposed to out-of-domain sounds or trained exclusively with P-Loss. Subjective hearing tests were conducted, showing a preference for the fine-tuned model~\cite{masuda2021soundmatch}. Masuda \textit{et al.} later extended this work with semi-supervised learning, highlighting significant gaps in in-domain and out-of-domain performance~\cite{masuda2023improving}.

Rather than focusing on a particular implementation, Shier \textit{et al.} presented Spieglib, a library for implementation of sound-matching pipelines~\cite{shier2020spiegelib}. This library provides different choices for DNNs, GAs, synthesizers, and feature extractors. Shier \textit{et al.} presented an experiment with a similar setup to Yee-King \textit{et al.}~\cite{yee2018automatic}, however, they found a genetic algorithm to be the best performing.

@@

A neural approximation for another part of the supervised sound-matching chain---this time the synthesizer---was proposed by Barkan \textit{et al.}~\cite{barkan2023inversynthII}. As they noted, without a differentiable synthesizer, ``model-based'' (or what we call supervised) approaches cannot directly compare $x_o$ to $x_t$, often opting for P-Loss, which may not correctly map the parameters of sound to the output audio~\cite{esling2019flow,han2023perceptual,masuda2023improving}. Given a synthesizer and a large dataset of sounds and corresponding parameters, a model which maps sounds to parameters, and another model which approximates the synthesizer can be trained, allowing the implementation of a loss function which combines P-Loss with STFT differences. This ``Inversynth II'' (IS2) approach yielded significant improvements to previous works which did not use the STFT approximations for loss~\cite{esling2019flow,barkan2019inversynth}.

Barkan \textit{et al.} attempted to  improve the IS2 model with Inference-Time
Finetuning (ITF). For ITF, the synthesizer approximation is frozen and the encoder is iteratively finetuned for the sample using gradient descent in order to minimize P-Loss, stopping at convergence or when maximum number of iterations has reached~\cite{barkan2023inversynthII}. Barkan \textit{et al.} report that the application of ITF generally worsened the spectral and manual performance metrics, perhaps due to the discrepancy between the true and approximated synthesizer or overfitting.

@@

Some recent works have opted for the use of audio embeddings as a similarity metric. In a recent work, Cherep \textit{et al.} used latent representations from the CLAP model~\cite{wu2023large} along with a differentiable synthesizer~\cite{synthhaxcherep2023} to create creative interpretations of sound effects. In their approach, a desired sound-effect is described in text and embedded using CLAP before a gradient free optimizer~\cite{evosax2022github} iteratively updates the synthesizer's parameters to minimize the difference between the text and output sound embeddings. Based on manual hearing tests, this approach did not produce the ``correct'' sounds more frequently than previous works~\cite{kreuk2022audiogen}. However, it did yield better scores for ``artistic-interpretation'' (or what this works refers to as ``imitation'').

@@

% \subsection{What is Lacking In the Field}
% \label{sec:lacking}
% %-----> much of the original text (available in older version) is removed for brevity)

% Having looked at the current available literature, we note four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
% \begin{enumerate}
%     \item \LossSelect: Is there a \textit{best performing} loss function, or is the selection of the loss function dependent on factors such as the sound domain, synthesizer capability and the desired characteristics of the output? The few existing works addressing this issue evaluate their proposed losses using only simplistic sound synthesis methods\todo{cite?}.
%     \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The use of different DSP functions for iterative sound-matching and their effects on the outcome or the interactions with the loss function have not been tested.
%     \item \PeriodicLoss: The sinusoidal nature of loss function landscapes is a problem that appears frequently\todo{cite?}. In differentiable settings, this problem causes gradient descent updates to local minima. 
%     \item \OutDomain: Sound-matching with out-of-domain sounds is an unexplored area, yet a necessary one for practical applications for sound designers.
% \end{enumerate}

% \subsection{What Approach Should We Take?}
% We argue that isolated benchmarking is a necessary precursor to building robust, generalizable sound-matching systems. To address the problems of \LossSelect{} and \SynthSelect{}, we pursue a methodology with 3 key characteristics that do not appear together in other works.

% 1. Simple differentiable synthesizers using common DSP functions, filters, and oscillators would be beneficial in addressing the \SynthSelect{} problem. Isolating small-scale, low-parameter synthesizers avoids the confounding effects of large neural proxies or embedding models, making it possible to directly analyze the interaction between synthesis structure and loss function.

% 2. Evaluation of experiments under multiple loss functions: Addressing the \LossSelect{} problem requires that multiple loss functions are tested. A comparison of performance of different loss functions is rare in previous works~\todo{cite}.

% 3. Parameters optimized directly with gradient descent without additional neural network approximations. This would complement prior neural and proxy-based methods by providing the missing “low-level experiments” that clarify fundamental interactions between DSP functions and similarity measures.

% These characteristics would fill a gap in prior work where there is little insight into the comparison of loss functions under different synthesis conditions, and how the two interact.

% % 4. Past synths are either simple differentiable functions (commonly FM), or complex VSTs where we can't isolate the interactions between loss and functions, here we use isolated building blocks of DSP functions in a differentiable environment.

>>>>

Masuda \textit{et al.} also applied supervised learning~\cite{masuda2021soundmatch}. Their work highlights the issue of non-linearity in parameter-to-synthesizer outputs and out-of-domain search. This work uses a differentiable subtractive synthesizer (two oscillators and an LP filter) alongside a NN model pre-trained with P-Loss on an in-domain dataset of randomly selected parameters. After training, the model was fine-tuned using 20,000 out-of-domain sounds from the NSynth dataset~\cite{engel2017neural} and multi-scale spectrogram loss~\cite{engel2020ddsp}. This approach proved more effective---i.e., achieved lower multi-level spectral difference in out-of-domain tests---than baseline models, which were either not exposed to out-of-domain sounds or trained exclusively with P-Loss. Subjective hearing tests were conducted, showing a preference for the fine-tuned model~\cite{masuda2022soundmatch}. Masuda \textit{et al.} later extended this work with semi-supervised learning, highlighting significant gaps in in-domain and out-of-domain performance~\cite{masuda2023improving}.

Rather than focusing on a particular implementation, Shier \textit{et al.} presented Spieglib, a library for implementation of sound-matching pipelines~\cite{shier2020spiegelib}. This library provides different choices for DNNs, GAs, synthesizers, and feature extractors. Shier \textit{et al.} presented an experiment with a similar setup to Yee-King \textit{et al.}~\cite{yee2018automatic}, however, they found a genetic algorithm as the best performing.

@@

A neural approximation for another part of the supervised sound-matching chain---this time the synthesizer---was proposed by Barkan \textit{et al.}~\cite{barkan2023inversynthII}. As they noted, without a differentiable synthesizer, ``model-based'' (or what we call supervised) approaches cannot directly compare $x_o$ to $x_t$, often opting for P-Loss, which may not correctly map the parameters of sound to the output audio~\cite{esling2019flow,han2023perceptual,masuda2023improving}. This approach used a model mapping sounds to parameters, another approximating the synthesizer, and a loss function which combines P-Loss with STFT differences. This ``Inversynth II'' (IS2) approach yielded significant improvements to previous works which did not use the STFT approximations for loss~\cite{esling2019flow,barkan2019inversynth}. Barkan \textit{et al.} then attempted to improve the IS2 model with Inference-Time
Fine-tuning (ITF). For ITF, the synthesizer approximation is frozen and the encoder is iteratively fine-tuned for a particular sample. However, ITF often degraded performance, possibly due to proxy–synthesizer mismatch or overfitting.

@@

 Audio embeddings can also be used as a similarity metric. In a recent work, Cherep \textit{et al.} used latent representations from the CLAP model~\cite{wu2023large} along with a differentiable synthesizer~\cite{synthhaxcherep2023} to create creative interpretations of sound effects. In their approach, a desired sound-effect is described in text and embedded using CLAP, followed by iterative updates from a gradient-free optimizer~\cite{evosax2022github} to the synthesizer's parameters to minimize embedding differences. Based on manual hearing tests, this approach did not produce the ``correct'' sounds more frequently than previous works~\cite{kreuk2022audiogen}. However, it did yield better scores for ``artistic-interpretation'' (or what this works refers to as ``imitation'').

@@

\subsection{What is Lacking In the Field}
\label{sec:lacking}
%-----> much of the original text (available in older version) is removed for brevity)

Having looked at the current available literature, we note four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
\begin{enumerate}
    \item \LossSelect: Is there a \textit{best performing} loss function, or is the selection of the loss function dependent on factors such as the sound domain, synthesizer capability and the desired characteristics of the output? The few existing works addressing this issue evaluate their proposed losses using only simplistic sound synthesis methods\todo{cite?}.
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The use of different DSP functions for iterative sound-matching and their effects on the outcome or the interactions with the loss function have not been tested.
    \item \PeriodicLoss: The sinusoidal nature of loss function landscapes is a problem that appears frequently\todo{cite?}. In differentiable settings, this problem causes gradient descent updates to local minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is an unexplored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

\subsection{What Approach Should We Take?}
We argue that isolated benchmarking is a necessary precursor to building robust, generalizable sound-matching systems. To address the problems of \LossSelect{} and \SynthSelect{}, we pursue a methodology with 3 key characteristics that do not appear together in other works.

1. Simple differentiable synthesizers using common DSP functions, filters, and oscillators would be beneficial in addressing the \SynthSelect{} problem. Isolating small-scale, low-parameter synthesizers avoids the confounding effects of large neural proxies or embedding models, making it possible to directly analyze the interaction between synthesis structure and loss function.

2. Evaluation of experiments under multiple loss functions: Addressing the \LossSelect{} problem requires that multiple loss functions are tested. A comparison of performance of different loss functions is rare in previous works~\todo{cite}.

3. Parameters optimized directly with gradient descent without additional neural network approximations. This would complement prior neural and proxy-based methods by providing the missing “low-level experiments” that clarify fundamental interactions between DSP functions and similarity measures.

These characteristics would fill a gap in prior work where there is little insight into the comparison of loss functions under different synthesis conditions, and how the two interact.

% 4. Past synths are either simple differentiable functions (commonly FM), or complex VSTs where we can't isolate the interactions between loss and functions, here we use isolated building blocks of DSP functions in a differentiable environment.
